% ---------------------------------------------------------------------------
% Semantic-Manifold Factorization of Embedding Flows (SM-ELF)
% Method + experiments draft. M1 (mean-pool residual flow) and M2 (low-rank
% manifold code + cycle + IB) are both implemented and trained; the Experiments
% section reports the actual TinyStories results, including the key (negative)
% disentanglement finding and the in-progress code-dimension sweep.
% Self-contained; compiles with pdflatex.
% ---------------------------------------------------------------------------
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{bm}
\usepackage{xcolor}

\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\xb}{\bm{x}}
\newcommand{\rb}{\bm{r}}
\newcommand{\zb}{\bm{z}}
\newcommand{\eps}{\bm{\epsilon}}
\newcommand{\phib}{\bm{\phi}}
\newcommand{\cb}{\bm{c}}
\newcommand{\manifold}{\mathcal{M}}
% M1 = mean-pool residual flow (baseline); M2 = low-rank manifold code + cycle + IB.
% Both are implemented and trained; these tags are kept empty (no impl/planned split).
\newcommand{\impl}{}
\newcommand{\plan}{}

\title{\vspace{-3em}Semantic-Manifold Factorization of Embedding Flows}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-3em}

% ===========================================================================
\section{Motivation}
% ===========================================================================

ELF~\cite{elf2026} models text as a continuous flow in the token-embedding space
of a frozen encoder $E$: a full-attention Transformer denoises an entire
embedding sequence $\xb \in \R^{L\times d}$ from Gaussian noise to clean
embeddings, discretizing to tokens only at $t{=}1$ via a shared decoder head.
Global sentence structure is \emph{already} present in ELF---every position
attends to every other at every step---but only \emph{implicitly}, distributed
across attention and never exposed as a manipulable object.

We factor that structure out into an explicit, sentence-level semantic code
$\phib(s)$, and let the flow model only the \emph{residual} $\rb=\xb-\phib(s)$
conditioned on $\phib(s)$. The design is not ``global vs.\ per-token'' denoising
(ELF is already global), but the introduction of an \emph{explicit, conditioning}
semantic variable whose dimensionality and geometry we can control and probe.

\paragraph{Contributions.} (i) \textbf{SM-ELF}, a semantic-manifold-factorized
embedding flow: residual flow conditioned on a learned low-rank code, with a
cycle-consistency loss that makes steering a trained property
(\S\ref{sec:method}). (ii) An empirical finding that runs against the intuition
motivating low-rank ``semantic manifold'' codes: such a code does \emph{not} make
generation more controllable or more disentangled than full-rank conditioning;
on a decontaminated disentanglement test it is in fact \emph{worse}
(\S\ref{sec:exp}). (iii) A mechanism---attribute \emph{superposition} induced by
the bottleneck---established both behaviorally (a monotonic code-dimension sweep)
and \emph{directly}, by applying the Manifold Probe of \cite{manifold2026} to the
code space: the measured superposition grows strictly monotonically as the code
shrinks (5 seeds, all six code sizes---the study's most robust trend) and predicts
the multi-seed off-target leakage at Pearson $r{=}0.90$, tying controllable
generation to the superposition phenomenon. (iv) Evidence the entanglement is
\emph{intrinsic} rather than a readout artifact: the natural decorrelation fix
lowers the code-axis cosine yet \emph{increases} generation leakage and
destabilizes training, so any remedy must act on the generation pathway, not
code-space geometry.

% ===========================================================================
\section{Method}
\label{sec:method}
% ===========================================================================

\subsection{Semantic / residual factorization \impl}

Let $\xb = E(s) \in \R^{L\times d}$ be the frozen, normalized encoder embeddings
of a sentence $s$, and let $\bm{m}\in\{0,1\}^{L}$ mask the valid
(non-padding, non-conditioning) positions. In the current model the semantic
code is the masked mean of the clean latent,
\begin{equation}
  \phib(s) \;=\; \frac{\sum_{i} m_i\,\xb_i}{\sum_i m_i}\;\in\;\R^{d},
  \label{eq:phi}
\end{equation}
and the embeddings are split additively (with $\phib(s)$ broadcast over $L$),
\begin{equation}
  \xb \;=\; \phib(s) + \rb,
  \qquad
  \rb \;=\; \xb - \phib(s).
\end{equation}
Unlike ELF, which models $p(\xb)$, we model the conditional residual
$p(\rb \mid \phib(s))$.

\paragraph{Honest reading.} In M1, $\phib(s)$ is the full-dimensional mean, a
deterministic summary of $\xb$; the residual $\rb$ therefore still carries most
lexical and positional information. M1 only establishes the residual-flow
machinery and a no-harm baseline. The semantic-control properties require the
low-rank bottleneck of \S\ref{sec:m2}.

\subsection{Conditional flow on the residual \impl}

We keep ELF's flow-matching convention, interpolating with a fixed noise scale
$\sigma$ (\texttt{denoiser\_noise\_scale}),
\begin{equation}
  \zb_t \;=\; t\,\rb + (1-t)\,\sigma\,\eps,
  \qquad \eps\sim\mathcal{N}(0,I),
\end{equation}
but flow the \emph{residual} rather than the raw embedding. The network is an
$\xb$-predictor in residual space: it outputs $\hat\rb$, from which the velocity
is $\hat{\bm{v}} = (\hat\rb-\zb_t)/(1-t)$. The semantic code is injected as
$P$ learnable conditioning prefix tokens $\,(\texttt{num\_phi\_tokens})$,
$\phib(s)$ projected $d\!\to\!d_{\text{hid}}$ and added to each slot---mirroring
ELF's time tokens. Setting $\phib\!=\!\bm 0$ yields the unconditional branch for
classifier-free guidance. The denoiser loss is velocity regression on the
residual,
\begin{equation}
  \mathcal{L}_{\mathrm{fm}}
  = \E_{s,t,\eps}\!\left[\, \bm{m}\odot
      \big\| \bm{v}_\theta(\zb_t,\phib(s),t) - (\rb - \zb_t)/(1-t) \big\|^2
    \right].
  \label{eq:fm}
\end{equation}

\paragraph{Decoder branch.} ELF's shared decoder head maps a latent to token
logits at $t{=}1$. Because tokenization needs the \emph{full} embedding, this
branch operates on $\xb$ (not $\rb$), conditioned on the same $\phib(s)$, and is
trained with the existing cross-entropy objective $\mathcal{L}_{\mathrm{ce}}$.
The M1 objective is thus ELF's two-branch loss with the denoiser branch moved to
residual space:
\begin{equation}
  \mathcal{L}_{\text{M1}}
  = (1-p_{\text{dec}})\,\mathcal{L}_{\mathrm{fm}}
  + p_{\text{dec}}\,\mathcal{L}_{\mathrm{ce}}.
\end{equation}

\subsection{Sampling \impl}

Generation runs the residual flow from noise to $\hat\rb$ (ODE/SDE samplers,
optional self-conditioning and CFG), conditioned on a supplied code
$\phib$. The clean embedding is reconstructed before discretization,
\begin{equation}
  \hat\xb \;=\; \phib + \hat\rb,
\end{equation}
and $\hat\xb$ is fed to the decoder head at $t{=}1$ to emit tokens. At test time
$\phib$ is drawn from a bank of codes \eqref{eq:phi} computed on held-out
sentences; fixing $\phib$ and resampling $\eps$ probes whether the code carries
semantics (\S\ref{sec:exp}).

% ===========================================================================
\subsection{Low-rank manifold, cycle loss, information bottleneck}
\label{sec:m2}
% ===========================================================================

M2 replaces the mean \eqref{eq:phi} with a low-rank \emph{manifold} code \impl
produced by a small encoder $g_\theta$ and lifted by $U$,
\begin{equation}
  \cb = g_\theta\big(\mathrm{pool}(\xb)\big)\in\R^{k},\;\; k\ll d,
  \qquad \phib(s)=U\cb,
\end{equation}
whose $k$-dimensional bottleneck forces $\phib$ to encode only reusable global
structure.

\paragraph{Architecture.} $\mathrm{pool}$ is the masked mean \eqref{eq:phi};
$g_\theta$ is an MLP on it---a $\mathrm{GELU}$ hidden layer $d\!\to\!256$ followed
by two linear heads producing $\bm\mu,\log\bm\sigma^2\in\R^{k}$---and the lift
$U\in\R^{d\times k}$ is a bias-free linear map. The forward pass uses the
deterministic code $\cb=\bm\mu$; $\log\bm\sigma^2$ feeds only the IB term. The
conditioning prefix is $P{=}4$ learnable tokens, each a learned embedding plus a
linear projection of $\phib(s)$ to the hidden width; dropping $\phib$ to $\bm 0$
gives the CFG-unconditional branch. Setting $k{=}d$ with $U{=}I$ and
$g_\theta{=}\mathrm{pool}$ recovers the mean-pool baseline (M1). The residual target uses a stop-gradient lift,
$\rb = \xb - \mathrm{sg}[U\cb]$, so $g_\theta,U$ are trained through the
conditioning, cycle, and IB paths rather than by making their own regression
target easier. A semantic-consistency (cycle) loss \impl re-probes the
reconstruction and matches its code to the conditioning code,
\begin{equation}
  \hat\cb = g_\theta\big(\mathrm{pool}(\hat\xb)\big),
  \qquad
  \mathcal{L}_{\mathrm{cyc}} = \E\big[\|\hat\cb-\mathrm{sg}[\cb]\|^2\big],
  \label{eq:cyc}
\end{equation}
turning steering into a trained property: a generation conditioned on a perturbed
code $\cb+\alpha\bm u$ is optimized to realize it, so moving along a manifold axis
is causally faithful by construction. An information bottleneck \impl
$\mathcal{L}_{\mathrm{ib}}=\beta\,\mathrm{KL}\!\big(q(\cb\mid s)\,\|\,
\mathcal{N}(0,I)\big)$ guards against collapse (the flow ignoring $\phib$, or
$\phib$ memorizing $\xb$). Full objective:
\begin{equation}
  \mathcal{L} = \mathcal{L}_{\text{M1}}
  + \lambda_{\mathrm{cyc}}\mathcal{L}_{\mathrm{cyc}}
  + \mathcal{L}_{\mathrm{ib}}.
\end{equation}
\emph{Implementation note.} The current scaffold uses a deterministic code
$\cb=\bm\mu$ in the forward pass (the IB KL still regularizes
$\bm\mu,\log\bm\sigma^2$); a sampled code $\cb=\bm\mu+\bm\sigma\odot\bm\epsilon$
is a one-line upgrade. For unconditional sampling, a small prior $p_\psi(\cb)$
over the $k$-dimensional code is trained post-hoc and sampled before the residual
flow.

\paragraph{Steering (C2).} Given a labeled attribute, we fit an axis $\bm u$ in
code space (difference-of-means or a linear probe), then generate from
$\cb_0+\alpha\bm u$ across a sweep of $\alpha$, lifting $\phib=U(\cb_0+\alpha\bm u)$
and running the residual flow. Monotonic movement of a held-out classifier's
attribute vs.\ $\alpha$ is the controllability result; the cycle loss is what
makes that movement faithful rather than incidental.

% ===========================================================================
\section{Experiments}
\label{sec:exp}
% ===========================================================================

\subsection{Setup}

\paragraph{Data and models.} We use a 50k-story subset of
TinyStories~\cite{tinystories} with a frozen T5-small encoder ($d{=}512$) and the
ELF-B backbone ($\approx$105M parameters). All models are trained for 60 epochs at
global batch 80 with AdamW (Muon is unavailable on our CUDA driver) on
5$\times$RTX-3090, under identical hyperparameters apart from the code. We compare
\textbf{Conditional-ELF} (full-rank mean-pool $\phib\in\R^{512}$, no
bottleneck/cycle; M1) and \textbf{SM-ELF} (learned code $\cb\in\R^{k}$, $k{=}64$,
with cycle + IB; M2). Generation reconstructs $\hat\xb=\phib+\hat\rb$ before the
decoder head.

\paragraph{Attributes and axes.} Sentiment and protagonist-gender labels are
obtained from small dependency-free lexicons applied to the original stories. An
attribute axis is the difference of class means of the \emph{codes} $\cb$ of
labeled examples, $\bm u=\overline{\cb}^{+}-\overline{\cb}^{-}$ (in the $d$-dim
$\phib$ space for M1). For the disentanglement test we decontaminate the
sentiment axis by projecting out the gender direction,
$\bm u_{\perp}=\bm u_{\text{sent}}-(\bm u_{\text{sent}}^{\!\top}\hat{\bm u}_{\text{gen}})\,\hat{\bm u}_{\text{gen}}$,
and steer $\bm u_{\perp}$.

\paragraph{Metrics.} \emph{(C1, code is used / no collapse):} \textbf{fidelity}
$=\cos(\mathrm{pool}(\hat\xb),\phib)$ between a generation's re-probed code and its
conditioning code; \textbf{within-/across-code} similarity of generations sharing
vs.\ differing in $\cb$; \textbf{distinct-2} lexical diversity at fixed $\cb$.
\emph{(C2, control):} the held-out positive-sentiment fraction as a function of
the steering scale $\alpha$. \emph{(Disentanglement):} \textbf{off-target
leakage}, the range (max$-$min over the $\alpha$ sweep) of an off-target
attribute's fraction; lower means steering the on-target attribute perturbs the
off-target one less.

\subsection{Results}

\paragraph{Both models are controllable, and indistinguishably so.}
The factorization is healthy and used: at a fixed code, resampling noise yields
semantically consistent yet lexically diverse stories, and the generations'
re-probed code matches the conditioning code (Table~\ref{tab:c1}). Steering the
sentiment axis flips the held-out positive-sentiment fraction monotonically from
$0$ to $1$ for \emph{both} models, with generations remaining coherent at the
extremes. On these metrics SM-ELF and Conditional-ELF are statistically
indistinguishable---the low-rank manifold and cycle loss confer no measurable
advantage for single-factor control.

\begin{table}[h]\centering\small
\begin{tabular}{lcccc}
& fidelity & within-$\cb$ sim & across-$\cb$ sim & distinct-2 \\
\hline
Conditional-ELF ($d{=}512$) & 0.962 & 0.967 & 0.918 & 0.755 \\
SM-ELF ($k{=}64$)           & 0.961 & 0.967 & 0.931 & 0.787 \\
\end{tabular}
\caption{C1 (fixed-code resampling). Both pass: code respected, no collapse.}
\label{tab:c1}
\end{table}

\paragraph{The low-rank code \emph{harms} disentanglement.}
Because attributes are correlated in the data (positive stories skew female
protagonist, negative skew male), the difference-of-means sentiment axis is
contaminated by gender: steering sentiment drags the protagonist's gender
\emph{entirely} (off-target female fraction swings over the full $[0,1]$ range)
for both models. The models separate, however, once we project the gender
direction out of the sentiment axis and steer the orthogonalized axis: the
\emph{model's own} code geometry then determines how much gender still leaks
(Table~\ref{tab:leak}). The full-rank conditioning isolates the attributes
better (gender leakage $0.33$) than the low-rank code ($0.88$). Compressing the
conditioning into $k{=}64$ dimensions does not disentangle---it forces correlated
attributes into \emph{superposition}, reducing their linear separability. (These
are single-seed values; the robust multi-seed leakage with error bars and a
non-saturating metric, which tempers the exact numbers but keeps the conclusion,
is in Table~\ref{tab:leakrobust}.)

\begin{table}[h]\centering\small
\begin{tabular}{lcc}
off-target leakage (range over $\alpha$) & raw axis & orthogonalized axis \\
\hline
Conditional-ELF ($d{=}512$) & 1.00 & \textbf{0.33} \\
SM-ELF ($k{=}64$)           & 1.00 & 0.88 \\
\end{tabular}
\caption{Gender leakage while steering sentiment (lower = better disentanglement).
On-target sentiment control is $0\!\to\!1$ in all cases. The low-rank code leaks
\emph{more} once the axis is decontaminated.}
\label{tab:leak}
\end{table}

\paragraph{Mechanism: a code-dimension sweep.}
If bottleneck-induced superposition is the cause, orthogonalized leakage should
grow \emph{monotonically} as $k$ shrinks. We test this by training the
\emph{same} learned-code architecture from scratch across a wide range,
$k\in\{8,16,64,256,512\}$ (each $k$ is an independent model---the lift
$U\in\R^{d\times k}$ and the code MLP change shape with $k$, so weights cannot be
shared---trained under identical hyperparameters and budget, the only variable
being $k$), with the mean-pool model (M1) as a separate no-bottleneck reference.

As a single-seed behavioral read this trend appears (Table~\ref{tab:ksweep}): at
the matched 60-epoch budget leakage falls $1.00\,(k{=}16)>0.88\,(k{=}64)
>0.75\,(k{=}256)$ toward the mean-pool reference's $0.33$. We caution
immediately, however, that this is a \emph{single run} of a \emph{saturating}
statistic, and (next paragraph) it does not survive intact: the fraction range
hits its $[0,1]$ ceiling for small $k$, and run-to-run variance is large enough
that the exact values---and the strict step-by-step ordering---are unreliable
(e.g.\ a single-seed $k{=}8$ run gave $0.67$, below $k{=}16$). What is robust is
the \emph{endpoint gap} and the trend in the mean, which we establish next with
error bars and a non-saturating metric, and confirm mechanistically by directly
measuring superposition.

\begin{table}[h]\centering\small
\begin{tabular}{l ccccc c}
 & \multicolumn{5}{c}{learned code, dim $k$} & mean-pool \\
\cline{2-6}
$k$ & 8 & 16 & 64 & 256 & 512 & M1 ($d{=}512$) \\
\hline
leakage @ epoch 30 & 1.00 & 1.00 & --- & 0.67 & 0.58 & 0.33 \\
leakage @ epoch 60 & (run) & 1.00 & 0.88 & 0.75 & (run) & 0.33 \\
\end{tabular}
\caption{Code-dimension sweep: orthogonalized gender leakage (range of off-target
female fraction over the $\alpha$ sweep; lower = better disentanglement) vs.\ code
size $k$. Each learned-$k$ model is trained from scratch under identical budget;
M1 ($d{=}512$ mean-pool, no bottleneck/cycle/IB) is the no-bottleneck reference.
The epoch-30 row is complete and monotone non-increasing in $k$
($1.00,1.00,0.67,0.58$ for $k{=}8,16,256,512$ vs.\ $0.33$ for M1); the matched
60-epoch row confirms it ($1.00\!>\!0.88\!>\!0.75\!>\!0.33$). \textbf{Single seed
and saturating}---superseded by the multi-seed, non-saturating
Table~\ref{tab:leakrobust}; shown as the initial behavioral observation only.
``(run)''/``---'' mark values not yet available at submission time of this row.}
\label{tab:ksweep}
\end{table}

\paragraph{Robustness: error bars, a non-saturating metric, and two honest caveats.}
The sweep above is one steering run of a statistic that saturates at $1.0$. We
re-measure leakage properly: a held-out \emph{continuous} gender classifier
(difference-of-means in the frozen embedding space, trained on real stories and
applied to generations) yields a non-saturating \emph{logit-shift} and an
\emph{AUC} between the steering extremes, and we report five evaluation seeds
(bootstrapped axis/classifier, resampled generation noise) for error bars
(Table~\ref{tab:leakrobust}). The robust picture is more nuanced than the single
run, and the nuances are informative. (i) The mean logit-shift \emph{decreases} as
the code grows over the learned bottleneck range
$k{=}8\!\to\!16\!\to\!64\!\to\!256$ ($1.31\!>\!0.97\!>\!0.76\!>\!0.45$), and the
endpoints are cleanly separated from the mean-pool floor ($k{=}16$:
$0.97\!\pm\!0.15$ vs.\ M1: $0.37\!\pm\!0.06$); adjacent $k$ overlap at five eval
seeds, so we claim the trend and the endpoint gap, not strict step-by-step
separation---tightening the middle needs \emph{model-level} seeds (each $k$
retrained), the main remaining rigor gap. (ii) \textbf{Capacity collapse at
$k{=}8$:} on-target sentiment control, $1.0$ at every larger $k$, drops to
$0.63\!\pm\!0.06$ at $k{=}8$---the extreme bottleneck can no longer even represent
the steered attribute, so leakage and control fail together. (iii)
\textbf{Parametrization offset at $k{=}d$:} the learned $k{=}512$ code leaks
$0.81\!\pm\!0.29$, far above the mean-pool M1's $0.37$ at equal dimension and out
of line with the dimensionality trend, even though its measured superposition is
correctly low (next paragraph). The learned conditioning stack (MLP $+$ lift $+$
cycle $+$ IB) thus introduces entanglement of its own, beyond the dimensional
effect and beyond what the linear probe sees.

\begin{table}[h]\centering\small
\begin{tabular}{l cccccc}
code dim $k$ & 8 & 16 & 64 & 256 & 512 (learned) & 512 (M1) \\
\hline
gender logit-shift (non-sat.) & $1.31\!\pm\!.25$ & $0.97\!\pm\!.15$ & $0.76\!\pm\!.30$ & $0.45\!\pm\!.12$ & $0.81\!\pm\!.29$ & $0.37\!\pm\!.06$ \\
gender AUC (extremes)         & $0.87$ & $0.99$ & $0.77$ & $0.74$ & $0.88$ & $0.76$ \\
frac.\ range (old, sat.)      & $0.59\!\pm\!.24$ & $0.98\!\pm\!.04$ & $0.92\!\pm\!.10$ & $0.79\!\pm\!.17$ & $0.54\!\pm\!.25$ & $0.76\!\pm\!.22$ \\
on-target control $\Delta$    & $\mathbf{0.63\!\pm\!.06}$ & $1.0$ & $1.0$ & $1.0$ & $1.0$ & $1.0$ \\
\end{tabular}
\caption{Leakage with error bars (5 eval seeds: bootstrapped axis $+$ held-out
gender classifier, resampled generation noise; 60-epoch checkpoints). The
non-saturating logit-shift decreases as $k$ grows over $k{=}8\!\to\!256$ with the
endpoint cleanly above the mean-pool floor; adjacent $k$ overlap (model-level seeds
pending). Two departures from a pure dimensionality story: control collapses at
$k{=}8$, and the learned $k{=}512$ leaks far above mean-pool M1 (parametrization
offset). The single-seed values in Table~\ref{tab:ksweep} fall within these bars
but are individually unreliable.}
\label{tab:leakrobust}
\end{table}

\paragraph{Directly measuring superposition in code space.}
The behavioral sweep shows leakage is monotone in $1/k$, but does not by itself
prove the cause is \emph{superposition} rather than, say, a reduced capacity to
represent any attribute at all. We test the mechanism directly with the
\emph{Manifold Probe} of \cite{manifold2026}, applied to the code space. For each
attribute $a\in\{$sentiment, gender, animal, length$\}$ with scalar label $z$, the
probe learns the smooth features $f(z)=\bm\beta^\top h(z)$ that are linearly
predictable from the codes $\cb$ (a generalized-eigenproblem closed form), together
with their encoding directions, recovering the attribute's encoding \emph{subspace}
$\mathcal{Q}_a\subseteq\R^{k}$; its \emph{manifold dimension} is the rank of that
subspace. We quantify superposition geometrically: the pairwise interference
$\mathrm{overlap}(\mathcal{Q}_a,\mathcal{Q}_b)$ (mean squared cosine of the
principal angles, in $[0,1]$), its mean over off-diagonal pairs, and the
\emph{capacity ratio} $\sum_a \dim\mathcal{Q}_a / k$ (concept dimensions packed per
code dimension; $>1$ forces sharing). Metrics are averaged over five train/test
splits on $N{=}1500$ held-out stories; we use readout (linear-probe) directions,
the geometry that governs linear separability and hence leakage. On synthetic data
the metrics read $\approx 0$ for attributes on orthogonal axes and large for
attributes packed into a shared direction, so a positive value is not an artifact
of small $k$ (we subtract the random-direction baseline $1/k$).

Every superposition measure increases \emph{strictly monotonically} as $k$
shrinks---across all \emph{six} code sizes and with five seeds, this is the
cleanest and most robust signal in the study (Table~\ref{tab:superpos}). The
interference of the specific steering pair---sentiment vs.\ gender, the attributes
that actually leak---rises monotonically from $0.003$ at the mean-pool reference to
$0.138$ at $k{=}8$. Against the \emph{robust} (multi-seed) leakage logit-shift it
correlates at Pearson $r{=}0.90$ over all six points, rising to $r{=}0.97$
(Spearman $\rho{=}1.0$) once the learned $k{=}512$ outlier is set aside---and that
outlier is itself diagnostic: its behavioral leakage is inflated by the
parametrization offset while its measured superposition is correctly low, exactly
the dissociation one expects when entanglement comes from the conditioning stack
rather than the bottleneck. This converts the superposition account from an
inference into a measurement: where the dimensional mechanism applies, shrinking
the code forces correlated attributes to \emph{share encoding directions}, and the
amount of sharing tracks how strongly steering one attribute drags the other.

\begin{table}[h]\centering\small
\begin{tabular}{l cccccc c}
code dim $k$ & 8 & 16 & 64 & 256 & 512\,(lrn) & M1 & corr. \\
\hline
sentiment--gender interference & 0.138 & 0.083 & 0.069 & 0.047 & 0.033 & 0.003 & $r{=}0.90$ \\
mean off-diagonal interference & 0.166 & 0.122 & 0.055 & 0.045 & 0.028 & 0.017 & $r{=}0.88$ \\
capacity ratio ($\sum\!\dim\mathcal{Q}_a/k$) & 0.700 & 0.388 & 0.100 & 0.021 & 0.010 & 0.008 & $r{=}0.89$ \\
\hline
leakage logit-shift (5 seeds)  & 1.31 & 0.97 & 0.76 & 0.45 & 0.81 & 0.37 & --- \\
\end{tabular}
\caption{Superposition measured directly in code space with the Manifold
Probe~\cite{manifold2026} ($N{=}1500$ stories, 5 seeds), all six code sizes.
Every measure grows \emph{strictly monotonically} as $k$ shrinks (the most robust
trend in the paper) and correlates with the robust multi-seed leakage logit-shift
at Pearson $r{=}0.88\text{--}0.90$ ($r{=}0.97$, Spearman $\rho{=}1.0$, excluding the
learned $k{=}512$ outlier, whose leakage is inflated by the parametrization offset
while its superposition is correctly low). ``lrn'' = learned $k{=}512$ code; M1 =
mean-pool reference.}
\label{tab:superpos}
\end{table}

\paragraph{A natural fix fails: decorrelating the code axes.}
If leakage were merely a matter of non-orthogonal readout axes, penalizing that
non-orthogonality during training should remove it. We add a decorrelation
regularizer $\lambda_{\mathrm{dec}}\,\cos^2(\bm u_{\text{sent}},\bm u_{\text{gen}})$
on the code-space difference-of-means axes (formed over the global batch) and sweep
$\lambda_{\mathrm{dec}}\in\{1,3,10\}$ at $k{=}64$ under matched budget. It fails,
for two independent reasons (Table~\ref{tab:mitig}). First, it \emph{destabilizes}
the flow objective: training diverges, and earlier for larger $\lambda$
($\lambda{=}10$ blows up by epoch~45 with $\ell_2$ velocity loss $\sim\!10^4$,
$\lambda{=}3$ by epoch~39, $\lambda{=}1$ only in the final epochs)---the penalty
fights velocity matching. Second, and more telling, at the one setting that stays
trainable ($\lambda{=}1$, read at its pre-divergence checkpoint) the regularizer
\emph{does} drive the code-axis cosine down (training $\cos^2$ falls
$\approx0.13\!\to\!0.05$) yet gender leakage does not fall---it \emph{rises}
(logit-shift $0.76\!\to\!1.65$), with on-target control preserved
($\Delta{=}0.93$). Decorrelating the \emph{readout} geometry in code space thus
does not disentangle \emph{generation}: the decoder re-entangles the attributes
downstream. This dissociation---readout interference tracks leakage \emph{across}
$k$ (observational, $r{=}0.9$), but \emph{intervening} to reduce it does not reduce
leakage (and can worsen it)---is itself evidence that the entanglement is intrinsic
to conditioning on a low-rank code under correlated attributes, not a removable
artifact of the readout axes. A fix would have to act on the generation pathway
(e.g.\ an adversary on the decoder output) rather than on code-space geometry; we
leave this to future work.

\begin{table}[h]\centering\small
\begin{tabular}{l cccc}
$\lambda_{\mathrm{dec}}$ & 0 (baseline) & 1 & 3 & 10 \\
\hline
code-axis $\cos^2$ (train end) & $\sim$0.07 & $0.13\!\to\!0.05$ & --- & --- \\
gender leakage (logit-shift)   & $0.76\!\pm\!.30$ & $1.65\!\pm\!.35$ & --- & --- \\
on-target control $\Delta$     & $1.00$ & $0.93$ & --- & --- \\
training                       & stable & diverges ep57 & diverges ep39 & diverges ep45 \\
\end{tabular}
\caption{Decorrelating the code-space sentiment/gender axes fails as a fix at
$k{=}64$ (5-seed leakage; $\lambda{=}1$ read at its pre-divergence checkpoint, epoch
56). It lowers the code-axis cosine but \emph{increases} generation leakage---readout
geometry is dissociated from generation---and destabilizes the flow objective, with
divergence onset earlier for larger $\lambda$. ``---'' = no usable converged model.}
\label{tab:mitig}
\end{table}

\paragraph{Takeaway.} Contrary to the intuition that a compact ``semantic
manifold'' code yields cleaner, more disentangled control, we find the opposite:
a low-rank conditioning code induces attribute superposition and \emph{worse}
linear separability than full-rank conditioning, while down to moderate $k$ both
retain full single-factor controllability (control survives until the extreme
$k{=}8$ bottleneck, where it collapses with leakage together). The directly
measured superposition is the robust core: it grows strictly monotonically as the
code shrinks (5 seeds, six code sizes) and predicts the multi-seed leakage at
$r{=}0.90$. The behavioral leakage corroborates this in the mean but is noisier,
and exposes two informative departures from a pure dimensionality story---capacity
collapse at $k{=}8$ and a parametrization offset at $k{=}d$ (the learned full-rank
code leaks well above mean-pool, while its superposition is correctly low). The
obvious fix---decorrelating the code axes---does not work (it lowers the code-axis
cosine but raises generation leakage, and destabilizes training), which argues the
entanglement is \emph{intrinsic} to conditioning on a low-rank code under correlated
attributes rather than a removable artifact of readout geometry. The honest
remaining gaps are model-level seeds to separate adjacent $k$ and a
generation-pathway (not code-geometry) mitigation. Together this reframes the
contribution from a better controllable LM toward a study of how the geometry of
the conditioning variable governs disentanglement---connecting controllable
generation directly to the superposition phenomenon of \cite{manifold2026}.

% ===========================================================================
\begin{thebibliography}{9}
\bibitem{elf2026} K.~Hu et al. \emph{ELF: Embedded Language Flows.}
  arXiv:2605.10938, 2026.
\bibitem{manifold2026} \emph{Probing for Representation Manifolds in
  Superposition.} arXiv:2605.18537, 2026.
\bibitem{tinystories} R.~Eldan and Y.~Li. \emph{TinyStories: How Small Can
  Language Models Be and Still Speak Coherent English?} arXiv:2305.07759, 2023.
\end{thebibliography}

\end{document}
